Faithful Variable Screening for\\ High-Dimensional Convex Regression

### 2015-05-18 tex/scam/additive_faithful.tex
a

For a general regression function, an additive approximation may result in a
relevant variable being incorrectly marked as irrelevant. Such
mistakes are inherent to the approximation and may persist even in
the population setting.  In this section we give
examples of this phenomenon, and then show how the convexity
assumption
changes the behavior of the additive approximation. We work with $C=[0,1]^p$ as the support of the distribution in this section but all of our results apply to general hypercubes. We begin
with a lemma that characterizes the components of the additive approximation under mild conditions.

@@

We note that $\E[ f(X) -\sum_{k'\neq k} f^*_{k'}(X_{k'}) | x_k] - \E f(X)$ has mean zero as a function of $x_k$, which shows that $f^*_k$'s are feasible.

\end{proof}

@@

We now show that under a general class of distributions which we
characterize below, convex multivariate functions are additively
faithful. To simplify presentation, we restrict our attention to densities bounded away from $0/\infty$, that is, $0 < \inf p(\mathbf{x}) \leq \sup p(\mathbf{x}) < \infty$.

@@

The boundary flatness condition intuitively states that two
conditional densities $p(\mathbf{x}_{-j} \given x_j)$ and
$p(\mathbf{x}_{-j} \given x_j')$ are similar when $x_j$ and $x_j'$ are
both close to the same boundary point. It is thus much more general
than product densities. Boundary flatness is a weak condition because
it affects only an $\epsilon$-small region around the boundary;
$p(\mathbf{x}_{-j} \given x_j)$ can take arbitrary shapes away from
the boundary. Boundary flatness also allows arbitrary correlation
structure between the variables (provided $p(\mathbf{x}) >0$). In
Section~\ref{sec:boundary_flat}, we give a detailed discussion of the
boundary flatness condition and show examples of boundary flat
densities. In particular, we show that any density supported on a
compact set can be approximated arbitrarily well by boundary flat
densities.

@@

\begin{theorem}
\label{thm:convex_faithful}
Let $p(\mathbf{x})$ be a density supported on $C=[0,1]^p$ and bounded away from $0/\infty$ that satisfies the boundary flatness property.

Suppose $f$ is a convex with a bounded second derivative on an open set containing $C$, then $f$ is additively faithful under $p(\mathbf{x})$.
\end{theorem}

@@

We differentiate with respect to $x_k$ at $x_k = 0,1$ under the integral. The detail necessary to verify the validity of this operation is technical and given in Section~\ref{sec:dominated_condition} of the supplementary material.

@@

\begin{example}
\label{ex:joint_density_flat}
Boundary flatness is satisfied if the joint density becomes flat at the boundary. To be precise, let $p(\mathbf{x})$ be a joint density bounded away from $0/\infty$ with a bounded second derivative.

Suppose also, for all $j$,
$$
\partial_{x_j} p(x_j,\mathbf{x}_{-j}) =
\partial_{x_j}^2 p(x_j, \mathbf{x}_{-j}) = 0\quad  \trm{at $x_j = 0, 1$,}.
$$
It is then straightforward to show boundary flatness. One can first verify that the derivatives of the marginal density $p(x_j)$ vanishes at $x_j = 0,1$ and then apply the quotient rule on $\frac{p(x_j,\mathbf{x}_{-j}}{p(x_j)}$ to show that $\partial_{x_j} p(\mathbf{x}_{-j} \given x_j) = \partial_{x_j}^2 p(\mathbf{x}_{-j} \given x_j) = 0$ at $x_j = 0,1$ as well. 
\end{example}

@@

\begin{example}
\label{ex:mixture_approx}
Suppose $p_\epsilon(\mathbf{x})$ is a bounded density over $[\epsilon, 1-\epsilon]^p$ for some $0 < \epsilon < 1/2$. Let $q(\mathbf{x})$ be an arbitrary boundary flat density over $[0,1]^p$ (one can take the uniform density for instance). Define a mixture $p_{\lambda, \epsilon}(\mathbf{x}) = \lambda q(\mathbf{x}) + (1-\lambda) p_\epsilon(\mathbf{x})$ where $0 < \lambda \leq 1$, then $p_{\lambda, \epsilon}(\mathbf{x})$ is boundary flat over $[0,1]^p$.

@@

Luckily, we can restore faithfulness by coupling the $f^*_k$'s with a set of univariate concave fits on the \emph{residual} $f - f^*$:
\begin{equation}
\label{eqn:concave_postprocess}
g^*_k = \arg\min \Big\{
   \E\Bigl( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k(X_k) \Bigr)^2
    \,:\, g_k \in \mh \mathcal{C}^1, \E g_k(X_k) = 0 
  \Big\}.
\end{equation}

\begin{theorem}
\label{thm:acdc_faithful}
Suppose $p(\mathbf{x})$ is a density on $C=[0,1]^p$ bounded away from
$0/\infty$ that satisfies the boundary flatness condition. Suppose
that $f$ is convex with a bounded second derivative on an open set
around $C$. Let $f^*_k$ and $g^*_k$ be as defined in equations
\eqref{eqn:convex_additive} and \eqref{eqn:concave_postprocess}, then
the $f^*_k$'s and the $g^*_k$'s are unique. Furthermore, $f^*_k = 0$
and $g^*_k = 0$ implies that $\partial_{x_k} f(\mathbf{x}) = 0$, that
is, $f$ does not depend on~$x_k$.
\end{theorem}

@@

We denote $h^*_k(x_k) = \E[ f(X) - \phi(X_{-k}) \given x_k]$. $f(\mathbf{x})$ and $\phi(\mathbf{x}_{-k})$ are both bounded and so $h^*_k(x_k)$ is bounded as well. Therefore, $h^*_k$ is square integrable and there exists a Fourier series $s_n(x_k)$ convergent to $h^*_k$ in $L_2$. Since $p(\mathbf{x})$ is bounded, $$\lim_{n \rightarrow \infty} \E \left( s_n(X_k) - h^*_k(X_k ) \right)^2 \rightarrow 0$$ as well.

@@

If we can show that $\E h^*_k(X_k)^2 = 0$, we would apply Lemma~\ref{cor:faithfulness_extension} and finish the proof. So let us suppose for sake of contradiction that $\E h^*_k(X_k)^2 > 0$.\\

Let $0 < \epsilon < 1$ be fixed and let $n$ be large enough such that $\E( s_n(X_k) - h^*_k(X_k))^2 \leq \epsilon \E h^*_k(X_k)^2 $.

Since $s_n(x_k)$ is twice-differentiable and has a second derivative bounded away from $-\infty$, there exist some positive scalar $\alpha$ such that $s_n(x_k) + \alpha(x_k^2 - m_k^2)$ has a non-negative second derivative and is thus convex.

>>>>

For a general regression function, an additive approximation may result in a
relevant variable being incorrectly marked as irrelevant. Such
mistakes are inherent in the approximation and may persist even in
the population setting.  In this section we give
examples of this phenomenon, and then show how the convexity
assumption
changes the behavior of the additive approximation. We work with $C=[0,1]^p$ as the support of the distribution in this section but all of our results apply to general hypercubes. We begin
with a lemma that characterizes the components of the additive approximation under mild conditions.

@@

We note that $\E[ f(X) -\sum_{k'\neq k} f^*_{k'}(X_{k'}) | x_k] - \E
f(X)$ has mean zero as a function of $x_k$, which shows that the $f^*_k$s
are feasible.
\end{proof}

@@

We now show that under a general class of distributions which we
characterize below, convex multivariate functions are additively
faithful. To simplify the presentation, we restrict our attention to
densities bounded away from $0/\infty$, that is, $0 < \inf
p(\mathbf{x}) \leq \sup p(\mathbf{x}) < \infty$.

@@

The boundary flatness condition intuitively states that two
conditional densities $p(\mathbf{x}_{-j} \given x_j)$ and
$p(\mathbf{x}_{-j} \given x_j')$ are similar when $x_j$ and $x_j'$ are
both close to the same boundary point. It is thus much more general
than requiring the density to be a product density. Boundary flatness
is a weak condition because it affects only an $\epsilon$-small region
around the boundary; $p(\mathbf{x}_{-j} \given x_j)$ can take
arbitrary shapes away from the boundary. Boundary flatness also allows
arbitrary correlation structure between the variables (provided
$p(\mathbf{x}) >0$). In Section~\ref{sec:boundary_flat}, we give a
detailed discussion of the boundary flatness condition and show
examples of boundary flat densities. In particular, we show that any
density supported on a compact set can be approximated arbitrarily
well by boundary flat densities.

@@

\begin{theorem}
\label{thm:convex_faithful}
Let $p(\mathbf{x})$ be a density supported on $C=[0,1]^p$ and bounded
away from $0/\infty$ that satisfies the boundary flatness property.  
Suppose $f$ is a convex function with bounded second derivatives on an
open set containing $C$. Then $f$ is additively faithful under $p(\mathbf{x})$.
\end{theorem}

@@

We differentiate with respect to $x_k$ at $x_k = 0,1$ under the
integral. The details necessary to verify the validity of this
operation are technical and given in
Section~\ref{sec:dominated_condition} of the supplementary material.

@@

\begin{example}
\label{ex:joint_density_flat}
Boundary flatness is satisfied if the joint density becomes flat at
the boundary. To be precise, let $p(\mathbf{x})$ be a joint density
bounded away from $0/\infty$ with a bounded second derivative.
Suppose also, for all $j$,
$$
\partial_{x_j} p(x_j,\mathbf{x}_{-j}) =
\partial_{x_j}^2 p(x_j, \mathbf{x}_{-j}) = 0\quad  \trm{at $x_j = 0, 1$}.
$$ It is then straightforward to show boundary flatness. One can first
verify that the derivatives of the marginal density $p(x_j)$ vanish
at $x_j = 0,1$ and then apply the quotient rule on
$\frac{p(x_j, \mathbf{\scriptstyle x}_{-j})}{p(x_j)}$ to show that $\partial_{x_j}
p(\mathbf{x}_{-j} \given x_j) = \partial_{x_j}^2 p(\mathbf{x}_{-j}
\given x_j) = 0$ at $x_j = 0,1$ as well.
\end{example}

@@

\begin{example}
\label{ex:mixture_approx}
Suppose $p_\epsilon(\mathbf{x})$ is a bounded density over $[\epsilon,
  1-\epsilon]^p$ for some $0 < \epsilon < 1/2$. Let $q(\mathbf{x})$ be
an arbitrary boundary flat density over $[0,1]^p$ (one can take the
uniform density for instance). Define a mixture $p_{\lambda,
  \epsilon}(\mathbf{x}) = \lambda q(\mathbf{x}) + (1-\lambda)
p_\epsilon(\mathbf{x})$ where $0 < \lambda \leq 1$; then $p_{\lambda,
  \epsilon}(\mathbf{x})$ is boundary flat over $[0,1]^p$.

@@

Luckily, we can restore faithfulness by coupling the $f^*_k$s with a set of univariate concave fits on the \emph{residual} $f - f^*$:
\begin{equation}
\label{eqn:concave_postprocess}
g^*_k = \arg\min \Big\{
   \E\Bigl( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k(X_k) \Bigr)^2
    \,:\, g_k \in \mh \mathcal{C}^1, \E g_k(X_k) = 0 
  \Big\}.
\end{equation}

\begin{theorem}
\label{thm:acdc_faithful}
Suppose $p(\mathbf{x})$ is a density on $C=[0,1]^p$ bounded away from
$0/\infty$ that satisfies the boundary flatness condition. Suppose
that $f$ is convex with a bounded second derivative on an open set
around $C$. Let $f^*_k$ and $g^*_k$ be as defined in equations
\eqref{eqn:convex_additive} and \eqref{eqn:concave_postprocess}, then
the $f^*_k$s and the $g^*_k$s are unique. Furthermore, $f^*_k = 0$
and $g^*_k = 0$ implies that $\partial_{x_k} f(\mathbf{x}) = 0$, that
is, $f$ does not depend on~$x_k$.
\end{theorem}

@@

We denote $h^*_k(x_k) = \E[ f(X) - \phi(X_{-k}) \given
  x_k]$.  Since $f(\mathbf{x})$ and $\phi(\mathbf{x}_{-k})$ are both bounded,
$h^*_k(x_k)$ is bounded as well. Therefore, $h^*_k$ is square
integrable and there exists a Fourier series $s_n(x_k)$ convergent to
$h^*_k$ in $L_2$. Since $p(\mathbf{x})$ is bounded, $$\lim_{n
  \rightarrow \infty} \E \left( s_n(X_k) - h^*_k(X_k ) \right)^2
\rightarrow 0$$ as well.

@@

If we can show that $\E h^*_k(X_k)^2 = 0$, we can apply
Lemma~\ref{cor:faithfulness_extension} and finish the proof. So let us
suppose for sake of contradiction that $\E h^*_k(X_k)^2 > 0$.

Let $0 < \epsilon < 1$ be fixed and let $n$ be large enough such that $\E( s_n(X_k) - h^*_k(X_k))^2 \leq \epsilon \E h^*_k(X_k)^2 $.
Since $s_n(x_k)$ is twice-differentiable and has a second derivative
bounded away from $-\infty$, there exists a positive scalar $\alpha$
such that $s_n(x_k) + \alpha(x_k^2 - m_k^2)$ has a non-negative second
derivative and is thus convex.
